---
title: "CVM in Heidelberg"
author: "Marcel Reinmuth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

Quick exploratory analysis of cigarette vending machine locations in Heidelberg, uncovering insights and patterns to better understand their spatial distribution.

# Setup

This section for reproducibility tbd.

## libraries

```{r import libs, message=F}
# main libraries
library(sf)
library(units)
library(tidyverse)
library(spatstat)
library(ggeffects)
library(tmap)
```

## Data download & preprocess

```{r read in data}

cvm <- st_read("cvm_data.gpkg", layer = "cvm", quiet=T)
schools <- st_read("cvm_data.gpkg", layer = "schools", quiet=T)
kindergartens <- st_read("cvm_data.gpkg", layer = "kindergarten", quiet=T)
hd_boundary <- st_read("cvm_data.gpkg", layer = "hd bound", quiet=T)

```

Data was preprocess in QGIS as of now.

# ESDA

Calculate distances between each school/kindergarten and CVM location. Aggregate the resulting matrix rowwise (by school/kindergarten) to get the minimum distance towards every distinct CVM.

```{r distances}

dist_mat_schools <- st_distance(schools, cvm)
dist_mat_kindergartens <- st_distance(kindergartens, cvm)

distinct_min_schools <- apply(dist_mat_schools, 1, min)
distinct_min_kindergartens <- apply(dist_mat_kindergartens, 1, min)

```



| Type         | Extreme | Distance                                     |
|--------------|---------|----------------------------------------------|
| kindergarte  | min     | `r min(distinct_min_kindergartens) |> round(3)`          |
| kindergarten | max     | `r max(distinct_min_kindergartens) |> round(3)` |
| school       | min     | `r min(distinct_min_schools) |> round(3)`          |
| school       | min     | `r max(distinct_min_schools) |> round(3)`                |


Furthest distance towards a CVM is ~ 2600 Meter.
We create a continuous dataframe with values from 0 to 2600 Meter and add cumulative counts for CVM within the respective distance to/from schools/kindergartens.


```{r data crunch}
breaks <- seq(0, 2600, by = 1)  # Adjust the 'by' value as per your requirement

dist_counts_schools <- table(cut(distinct_min_schools, breaks = breaks, right = FALSE))
dist_counts_kindergartens <- table(cut(distinct_min_kindergartens, breaks = breaks, right = FALSE))

result_df <- data.frame(distance = breaks[2:length(breaks)],
                        cvm_count_schools = as.numeric(dist_counts_schools),
                        cvm_count_kindergartens = as.numeric(dist_counts_kindergartens))

result_df <- result_df |>
  mutate(cvm_count_schools_cum = cumsum(cvm_count_schools),
         cvm_count_kindergartens_cum = cumsum(cvm_count_kindergartens))

```


# Viz



```{r simple plot}
plot(result_df$distance, result_df$cvm_count_kindergartens_cum, type = "l", col = "blue", lwd = 2,
     xlab = "Distance in m", ylab = "cumulative count", main = "CVM distance to schools/kindergartens")
lines(result_df$distance, result_df$cvm_count_schools_cum, col = "green", lwd = 2)
legend("topleft", legend = c("kindergartens", "schools"), col = c("blue", "green"), lwd = 2, lty = 1)


# zoom 0 - 150

plot(result_df$distance[1:150], result_df$cvm_count_kindergartens_cum[1:150], type = "l", col = "blue", lwd = 2,
     xlab = "Distance in m", ylab = "cumulative count", main = "CVM distance to schools/kindergartens")
lines(result_df$distance[1:150], result_df$cvm_count_schools_cum[1:150], col = "green", lwd = 2)
legend("topleft", legend = c("kindergartens", "schools"), col = c("blue", "green"), lwd = 2, lty = 1)


```

The first plot indicates a steep growth in CVM locations within 0 - 500 meters distance to/from schools & kindergartens. However when zoomed in we see less than 10 schools and less than 30 kindergartens are within the first 100m.

But how many are exactly within 100m?

```{r cutoff100}

result_df |> 
  filter(distance<=100) |> 
  summarize(schools=sum(cvm_count_schools),
            kindergartens=sum(cvm_count_kindergartens))

```



```{r map, fig.width=6}
tmap_mode("view")

facilities <- rbind(schools |> select(c(amenity)),
                 kindergartens |> select(c(amenity)))

facilities_buff100 <- facilities |> st_transform(25832) |> st_buffer(100) |> st_transform(4326)

facilities_buff200 <- facilities |> st_transform(25832) |> st_buffer(200) |> st_transform(4326)

tm_basemap("OpenStreetMap") +
  tm_shape(facilities) +
  tm_symbols(
    col = "amenity",
    palette = c("darkgreen", "darkred"),
    size = 0.1,
    border.lwd = 0
  ) +
  tm_shape(cvm) +
  tm_symbols(col = "black", size = 0.1) +
  tm_shape(facilities_buff100) +
  tm_borders(col="grey50") +
  tm_shape(facilities_buff200) +
  tm_borders(col="grey25")

```



# Model

```{r poisson regression}

schools_m <- glm(cvm_count_schools_cum ~ distance, data = result_df, family = poisson)
summary(schools_m)
ggpredict(schools_m)

kindergarten_m <- glm(cvm_count_kindergartens_cum ~ distance, data = result_df, family = poisson)
summary(kindergarten_m)
ggpredict(kindergarten_m)

```


```{r point pattern}

cvm_ppp <- cvm |> st_transform(25832) |> as.ppp()
marks(cvm_ppp) <- NULL

hd_owin <- hd_boundary |> st_transform(25832) |> as.owin()
Window(cvm_ppp) <- hd_owin

ripley_K <- Kest(cvm_ppp, correction = "Ripley")
plot(ripley_K, main = "Ripley's K Function")
```

The K function indicates the location of CVMs are not subject to complete spatial randomness, but are clustered.

